Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-06-10 alignpair_letter.tex
Accepted Version

% Figures are 86 mm or 178 mm wide
\begin{figure}[h!]
    \centering%
    \includegraphics[scale=1]{figures/fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
        %
        Indels in protein-coding sequences can be classified as having one of three different phases and being one of two different types.
        Phases refer to the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only). The difference between an in-frame and a frameshift indel is not displayed.
        }
    \label{fig:aln}
\end{figure}

@@

COATi FST is not a true probabilistic FST \citep{cotterell-etal-2014-stochastic} and cannot be used as-is to simulate output sequences based on an input sequence. This is because it is missing a parameter to control how often Ns are added to the output sequence. This is a design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol. In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment. This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C. While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and eliding any states in between:
\[
\frac{1-e + g(e-g)^{n+1}}{1-e+g} \approx \frac{1-e}{1-e+g} 
\]
where $g$ is the gap opening parameter, $e$ is the gap-extension parameter, and $n$ is the length of the input sequence. The exponent has a $+1$ due to the presence of an ``immortal match'' immediately after the input sequence.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, ClustalΩ, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

@@

As expected, COATi was not able to align 23 sequence pairs and COATi-rev was not able to align 193 sequence pairs. PRANK was not able to align 24 sequence pairs. MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from its wrapper script. Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.

@@

In addition to supporting all three phases of gaps, COATi's advantage over the other aligners was at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}. This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.

@@

ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts. Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment. ClustalΩ was the worst performing protocol that we tested. It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.  --- It even performed poorly on its own gap patterns. --- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.

>>>>

% Figures are 86 mm or 178 mm wide
\begin{figure}[h!]
    \centering%
    \includegraphics[scale=1]{figures/fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows the true alignment of an ancestral sequence (A) and a descendant sequence (D).
        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
        %
        Indels in protein-coding sequences can be classified as having one of three different phases and being one of two different types.
        Phases refer to the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only). The difference between an in-frame and a frameshift indel is not displayed.
        }
    \label{fig:aln}
\end{figure}

@@

COATi FST is not a true probabilistic FST \citep{cotterell-etal-2014-stochastic} and cannot be used as-is to simulate output sequences based on an input sequence. This is because it is missing a parameter to control how often Ns are added to the output sequence. This is a design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol. In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment. This allows COATi FST to assign the same weight to gaps at the beginning and the end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C. While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and eliding any states in between:
\[
\frac{1-e + g(e-g)^{n+1}}{1-e+g} \approx \frac{1-e}{1-e+g} 
\]
where $g$ is the gap opening parameter, $e$ is the gap-extension parameter, and $n$ is the length of the input sequence. The exponent has a $+1$ due to the presence of an ``immortal match'' immediately after the input sequence.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, ClustalΩ, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments. We preserved the phases of gaps and the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

@@

As expected, COATi was not able to align 23 sequence pairs and COATi-rev was not able to align 193 sequence pairs. PRANK was not able to align 24 sequence pairs. MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from its wrapper script. Additionally, MACSE produced alignments with columns that contained only gaps, which were removed by its wrapper script.
A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.

@@

In addition to supporting all three phases of gaps, COATi's advantage over the other aligners was at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings. Different datasets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}. This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.

@@

ClustalΩ's amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts. Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment. ClustalΩ was the worst performing protocol that we tested. It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.  --- It even performed poorly on its own gap patterns. --- ClustalΩ's relatively poor performance likely derives from its default assumptions being a poor fit for our dataset and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.

### 2024-06-10 alignpair_letter_source.tex
Accepted Version

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, Clustal\textOmega{}, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

@@

As expected, COATi was not able to align 23 sequence pairs and COATi-rev was not able to align 193 sequence pairs. PRANK was not able to align 24 sequence pairs. MAFFT, MACSE, and Clustal\textOmega{} were able to align all sequence pairs, although the latter received some help from its wrapper script. Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.

@@

Clustal\textOmega{}'s amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts. Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment. Clustal\textOmega{} was the worst performing protocol that we tested. It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.  --- It even performed poorly on its own gap patterns. --- Clustal\textOmega{}'s relatively poor performance likely derives from its default assumptions being a poor fit for our data set and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.

@@

% Figures are 86 mm or 178 mm wide
\begin{figure}[ht]
    \centering%
    %\includegraphics[scale=1]{figures/fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
        %
        Indels in protein-coding sequences can be classified as having one of three different phases and being one of two different types.
        Phases refer to the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only). The difference between an in-frame and a frameshift indel is not displayed.
        \label{fig:aln}}
\end{figure}

>>>>

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, Clustal\textOmega{}, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments. We preserved the phases of gaps and the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

@@

As expected, COATi was not able to align 23 sequence pairs and COATi-rev was not able to align 193 sequence pairs. PRANK was not able to align 24 sequence pairs. MAFFT, MACSE, and Clustal\textOmega{} were able to align all sequence pairs, although the latter received some help from its wrapper script. Additionally, MACSE produced alignments with columns that contained only gaps, which were removed by its wrapper script.
A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.

@@

Clustal\textOmega{}'s amino-acid model also only permits phase-3 gaps and is very sensitive to frameshift artifacts. Because this approach depends on amino-acid translations, if a coding sequence contains an abiological frameshift, then translation will produce an abiological amino-acid sequence resulting in a poor alignment. Clustal\textOmega{} was the worst performing protocol that we tested. It had the highest average alignment error and also had difficulties correctly identifying positive selection and estimating evolutionary distances.  --- It even performed poorly on its own gap patterns. --- Clustal\textOmega{}'s relatively poor performance likely derives from its default assumptions being a poor fit for our dataset and also from the loss of information that occurred when we translated protein-coding DNA sequences into protein sequences.

@@

% Figures are 86 mm or 178 mm wide
\begin{figure}[ht]
    \centering%
    %\includegraphics[scale=1]{figures/fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows the true alignment of an ancestral sequence (A) and a descendant sequence (D).
        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
        %
        Indels in protein-coding sequences can be classified as having one of three different phases and being one of two different types.
        Phases refer to the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only). The difference between an in-frame and a frameshift indel is not displayed.
        \label{fig:aln}}
\end{figure}
